---
title: "SEM_spatmem"
author: "Zhiyi Sun"
date: "2023-12-30"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lavaan)
library(dplyr)
```

```{r}
data <- read.csv('spatmem_participant.csv')

parti <- c('Alex', 'Dorien', 'Fraukje', 'Frederike', 'Kanzi', 'Lobo', 'Lome', 'Teco')

data_cb <- data %>% filter(participant %in% parti)
data_human <- data %>% filter(!participant %in% parti)

data_cb <- data_cb[, -which(names(data_cb) %in% c("path", "hint.state", "start.position", "start.orientation", "start.treeSight"))] %>% filter(complete.cases(.)) 
data_human <- data_human[, -which(names(data_human) %in% c("path", "hint.state", "start.position", "start.orientation", "start.treeSight"))] %>% filter(complete.cases(.)) 
```

9. Structural Equation Modeling (SEM) (cannot be used to human group)

SEM can be used to analyze complex relationships between observed and latent variables, which is useful if you hypothesize a certain structure or causal relationship in your data.

Hypothesis: There is a structured relationship between variables that influence participant performance.

```{r}
stages <- c('1A', '1B', '1C', '2', '2B', '2A', '3A', '3B', '3C', '4', '4N')

# SEM model
sem_model <- '
  latent_variable =~ straightness.index + trialDuration
  latent_variable ~ changepoints
'
# Create a combined group identifier
# data_cb$combinedGroup <- interaction(as.factor(data_cb$studyGroup), as.factor(data_cb$stage.true))

for (i in stages) {
  temp <- subset(data_cb, stage.true == i)
  fit <- sem(sem_model, data = data_cb, group = "studyGroup")
  print(summary(fit))
}

# # SEM model
# sem_model <- '
#   latent_variable =~ straightness.index + trialDuration
#   latent_variable ~ changepoints
# '
# # Create a combined group identifier
# data_human$combinedGroup <- interaction(as.factor(data_human$studyGroup), as.factor(data_human$stage.true))
# 
# for (i in stages) {
#   temp <- subset(data_human, stage.true == i)
#   fit <- sem(sem_model, data = data_human, group = "studyGroup")
#   print(summary(fit))
# }
```

15. Structural Equation Modeling (SEM) with confounders (cannot be used to human group)

SEM is a comprehensive statistical technique that allows for the analysis of complex relationships between observed and latent variables.

Hypothesis: There are underlying latent variables that explain the relationships between observed performance metrics.

```{r}
# SEM model
sem_model <- '
  latent_variable_1 =~ straightness.index + trialDuration
  latent_variable_2 =~ trial_count + interactions
  latent_variable_1 ~ latent_variable_2 + changepoints
'
# Create a combined group identifier
# data_cb$combinedGroup <- interaction(as.factor(data_cb$studyGroup), as.factor(data_cb$stage.true))
 
for (i in stages) {
  temp <- subset(data_cb, stage.true == i)
  fit <- sem(sem_model, data = data_cb, group = "studyGroup")
  print(summary(fit))
}

# # SEM model
# sem_model <- '
#  latent_variable_1 =~ straightness.index + trialDuration
#  latent_variable_2 =~ trial_count + interactions
#  latent_variable_1 ~ latent_variable_2 + changepoints
# '
# # Create a combined group identifier
# data_human$combinedGroup <- interaction(as.factor(data_human$studyGroup), as.factor(data_human$stage.true))
#  
# for (i in stages) {
#   temp <- subset(data_human, stage.true == i)
#   fit <- sem(sem_model, data = data_human, group = "studyGroup")
#   print(summary(fit))
# }
```

Interpretation of Further Analyses:

9. SEM: Reveals complex causal relationships and the structure of interactions between variables, offering a comprehensive view of the factors influencing participant performance.
15. SEM: Provides a framework to understand the causal relationships and latent structures in the data.
