---
title: "spatmem_stage_1a"
author: "Zhiyi Sun"
date: "2023-08-07"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("emmeans")
library("dplyr") 
library("dslabs") 
library("ggplot2") 
library("ggrepel") 
library("scales")
library("ggridges")
library("tidyr") 
library("stringr") 
library("skimr")
library("MASS") 
library("naniar") 
library("rpart") 
library("simputation")
library("car")
library("DALEX")
library("superheat")
library("tibble")
library("ggfortify")
library("corrplot")
library("factoextra")
library("caret")
library("gridExtra")
library("patchwork")
library("glmnet")
library("ggpubr")
library("ggpmisc")
library("cluster")
library("lfda")
library("palmerpenguins")
library("ggh4x")
library("autoReg")
library("sjPlot")
library("sjmisc")
library("sjlabelled")
```

```{r}
load("stageSeparated 1.RData")
```

```{r}
dataset_SpatMem_1a <- rbind(kde.data.bantuchild.1a, kde.data.bonobo.1a, kde.data.chimp.1a, kde.data.mbendjelechild.1a, kde.data.usadult.1a, kde.data.uschild.1a)
```

# Check NAs

```{r}
colSums(is.na(dataset_SpatMem_1a))
```

# Str and table of studyGroup
```{r}
str(dataset_SpatMem_1a)

dataset_SpatMem_1a$targetCollected <- as.numeric(dataset_SpatMem_1a$targetCollected)
dataset_SpatMem_1a$straightness.index <- as.numeric(dataset_SpatMem_1a$straightness.index)
dataset_SpatMem_1a$changepoints <- as.numeric(dataset_SpatMem_1a$changepoints)
dataset_SpatMem_1a$trialDuration <- as.numeric(dataset_SpatMem_1a$trialDuration)
dataset_SpatMem_1a$interactions <- as.numeric(dataset_SpatMem_1a$interactions)
dataset_SpatMem_1a$trial_count <- as.numeric(dataset_SpatMem_1a$trial_count)

summary(dataset_SpatMem_1a)

table(dataset_SpatMem_1a$studyGroup)
```

# Distribution of targetCollected/straightness.index/changepoints/trialDuration/interactions/studyGroup
```{r}
par(mfrow = c(3,2))
barplot(table(dataset_SpatMem_1a$targetCollected),main="Distribution of targetCollected",xlab="targetCollected",ylab="Frequency")
hist(dataset_SpatMem_1a$straightness.index, breaks=50,main="Distribution of variable straightness.index",xlab="Number of straightness.index",ylab="Frequency")
hist(dataset_SpatMem_1a$changepoints, breaks=15,main="Distribution of variable changepoints",xlab="Number of changepoints",ylab="Frequency")
hist(dataset_SpatMem_1a$trialDuration, breaks=50,main="Distribution of variable trialDuration",xlab="Number of votes for trialDuration",ylab="Frequency")
hist(dataset_SpatMem_1a$interactions, breaks=50,main="Distribution of variable interactions",xlab="Number of votes for interactions",ylab="Frequency")
barplot(table(dataset_SpatMem_1a$studyGroup),main="Distribution of studyGroup",xlab="studyGroup",ylab="Frequency")
```

# Boxplot of changepoints across each studyGroup 
```{r}
boxplot(changepoints~studyGroup, data = dataset_SpatMem_1a, xlab="Study Group", ylab="Change Points", col = c("gold", "darkgreen"), cex.axis = 0.8)
```

# Hist facet
```{r}
agg <- aggregate(changepoints ~ studyGroup, dataset_SpatMem_1a, function(x) {
  c(mean = mean(x), median = median(x))
})
agg <- cbind(agg[1], agg[[2]])
agg <- reshape2::melt(agg, id.vars = "studyGroup")
names(agg)[names(agg) == 'variable'] <- 'stat'

ggplot(data = dataset_SpatMem_1a, aes(x = changepoints, fill=studyGroup)) + 
  geom_histogram(binwidth = 1) + 
  stat_theodensity(aes(y = after_stat(count))) +
  geom_vline(data = agg, mapping = aes(xintercept = value,
                                       color = stat)) +
  facet_wrap(~studyGroup)

ggplot(dataset_SpatMem_1a, aes(x = changepoints)) +
  geom_histogram(bins = 13) +
  geom_vline(data = agg, mapping = aes(xintercept = value,
                                       color = stat)) +
  facet_wrap(~studyGroup) +
  theme_bw()
```

# Fit a univariate linear regression - straightness.index/trialDuration ~ changepoints    
```{r}
fit1 <- lm(straightness.index~changepoints, data = dataset_SpatMem_1a)
summary(fit1)
tab_model(fit1)

fit2 <- lm(trialDuration~changepoints, data = dataset_SpatMem_1a)
summary(fit2)
tab_model(fit2)

tab_model(
  fit1, fit2, 
  dv.labels = c("straightness.index", "trialDuration"),
  string.pred = "Coeffcient",
  string.ci = "Conf. Int (95%)",
  string.p = "P-Value"
)
```

# Use t.test to compare straightness.index/trialDuration and changepoints
```{r}
t.test(dataset_SpatMem_1a$straightness.index, dataset_SpatMem_1a$changepoints)
t.test(dataset_SpatMem_1a$trialDuration, dataset_SpatMem_1a$changepoints)
```

# Summary Table
```{r}
summary(lm(changepoints~factor(studyGroup),data=dataset_SpatMem_1a))
tab_model(lm(changepoints~factor(studyGroup),data=dataset_SpatMem_1a),
  string.pred = "Coeffcient",
  string.ci = "95% CI",
  string.p = "P-Value")

tab_model(lm(straightness.index~factor(studyGroup),data=dataset_SpatMem_1a),
          lm(trialDuration~factor(studyGroup),data=dataset_SpatMem_1a),
  string.pred = "Coeffcient",
  string.ci = "95% CI",
  string.p = "P-Value")
```

# Fit the multiple linear regression - straightness.index
```{r}
fit2=lm(straightness.index~factor(studyGroup)*changepoints,data=dataset_SpatMem_1a)
summary(fit2)

fit3=lm(trialDuration~factor(studyGroup)*changepoints,data=dataset_SpatMem_1a)
summary(fit3)

tab_model(fit2,fit3,
  string.pred = "Coeffcient",
  string.ci = "95% CI",
  string.p = "P-Value")
```

# Forest
```{r}
plot_models(
  fit2, fit3,
  axis.labels = c(
    "US_Child × changepoints", "US_Adult × changepoints", "Mbendjele_Child × changepoints", "chimp × changepoints", "bonobo × changepoints", "changepoints", "US_Child", "US_Adult", "Mbendjele_Child", "chimp", "bonobo"),
  show.values = FALSE, show.p = FALSE, p.shape = TRUE, grid = TRUE
)

plot_models(
  fit2, fit3,
  axis.labels = c(
    "US_Child × changepoints", "US_Adult × changepoints", "Mbendjele_Child × changepoints", "chimp × changepoints", "bonobo × changepoints", "changepoints", "US_Child", "US_Adult", "Mbendjele_Child", "chimp", "bonobo"),
  show.values = FALSE, show.p = FALSE, p.shape = TRUE
)
```

# Create a plot of straightness.index versus changepoints for each level of studyGroup with confidence intervals at changepoints=0, 5, 6.368, 13
```{r}
fib.lm <- lm(straightness.index ~ changepoints*factor(studyGroup), data = dataset_SpatMem_1a)
fib.rg <- ref_grid(fib.lm, at = list(changepoints = c(0, 5, 6.368, 13)))
emmip(fib.rg, factor(studyGroup)~changepoints, CI=TRUE) 
```

# Create a plot of trialDuration versus changepoints for each level of studyGroup with confidence intervals at changepoints=0, 5, 6.368, 13
```{r}
fib.lm <- lm(trialDuration ~ changepoints*factor(studyGroup), data = dataset_SpatMem_1a)
fib.rg <- ref_grid(fib.lm, at = list(changepoints = c(0, 5, 6.368, 13)))
emmip(fib.rg, factor(studyGroup)~changepoints, CI=TRUE) 
```

# Make a plot that shows the relationship between the category of studyGroup and changepoints
```{r}
ggplot(dataset_SpatMem_1a, aes(x = studyGroup, y = changepoints)) +
geom_jitter() +
scale_x_discrete() +
theme(axis.text.x = element_text(angle = 90)) + 
labs(x = "studyGroup", y = "changepoints")

## categories are sorted from those with most to least average changepoints
ggplot(dataset_SpatMem_1a, aes(x = reorder(studyGroup, -changepoints, mean), y = changepoints)) + 
geom_jitter() +
scale_x_discrete() +
theme(axis.text.x = element_text(angle = 90)) + 
labs(x = "studyGroup", y = "changepoints")
```

## Take a look at targetCollected = 1
```{r}
z = subset(dataset_SpatMem_1a, targetCollected == 1, select = logFileName:trial_count)
ggplot(dataset_SpatMem_1a, aes(x = reorder(studyGroup, -changepoints, mean), y = changepoints, color = as.factor(targetCollected))) +
geom_jitter() +
scale_x_discrete() +
scale_color_brewer(palette = "Set2") + 
geom_text_repel(data=z, aes(x = studyGroup, y = changepoints, label=participant, colour = as.factor(targetCollected)), segment.size = 0.2) +
theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + 
labs(x = "studyGroup", y = "changepoints", color = "targetCollected")
```

# Showing the changepoints according to different studyGroup - straightness.index/trialDuration
```{r}
ggplot(dataset_SpatMem_1a, aes(x = straightness.index, y = changepoints, colour = studyGroup)) + 
geom_line(aes(group = targetCollected), linewidth = 1) + 
scale_color_brewer(palette = "Set2") +
facet_wrap(~ studyGroup, ncol=2) +
labs(x = "straightness.index", y = "changepoints") + 
theme_minimal()
```

```{r}
ggplot(dataset_SpatMem_1a, aes(x = trialDuration, y = changepoints, colour = studyGroup)) + 
geom_line(aes(group = targetCollected), linewidth = 1) + 
scale_color_brewer(palette = "Set2") +
facet_wrap(~ studyGroup, ncol=2) +
labs(x = "trialDuration", y = "changepoints") + 
theme_minimal()
```

# Showing the changepoints according to different studyGroup - lm-straightness.index/trialDuration
```{r}
ggplot(dataset_SpatMem_1a, aes(x = straightness.index, y = changepoints, colour = studyGroup)) + 
geom_point(size=0.5) + 
geom_smooth(method = lm) +
scale_color_brewer(palette = "Set2") +
facet_wrap(~ studyGroup, ncol=3) +
labs(x = "straightness.index", y = "changepoints") + 
stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), label.y = 13, size = 3) + 
theme_minimal()
```

```{r}
ggplot(dataset_SpatMem_1a, aes(x = trialDuration, y = changepoints, colour = studyGroup)) + 
geom_point(size=0.5) + 
geom_smooth(method = lm) +
scale_color_brewer(palette = "Set2") +
facet_wrap(~ studyGroup, ncol=3) +
labs(x = "trialDuration", y = "changepoints") + 
stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), label.y = 13, size = 3) + 
xlim(50, 150) + 
theme_minimal()
```

# Relationship between straightness.index/trialDuration and changepoints
```{r}
ggplot(dataset_SpatMem_1a, aes(x = straightness.index, y = changepoints, colour = studyGroup)) + 
geom_point() +
scale_color_brewer(palette = "Set2") +
facet_wrap(~ targetCollected, ncol=6) + 
theme_minimal() + 
theme(legend.position = "bottom") + 
labs(x = "straightness.index", y = "changepoints")

ggplot(dataset_SpatMem_1a, aes(x = straightness.index, y = changepoints, colour = as.factor(targetCollected))) + 
geom_point() +
scale_color_brewer(palette = "Set2") +
facet_wrap(~ studyGroup, ncol=6) + 
theme_minimal() + 
theme(legend.position = "bottom") + 
labs(x = "straightness.index", y = "changepoints", colour = "targetCollected")
```

```{r}
ggplot(dataset_SpatMem_1a, aes(x = trialDuration, y = changepoints, colour = studyGroup)) + 
geom_point() +
scale_color_brewer(palette = "Set2") +
facet_wrap(~ targetCollected, ncol=6) + 
theme_minimal() + 
theme(legend.position = "bottom") + 
labs(x = "trialDuration", y = "changepoints")

ggplot(dataset_SpatMem_1a, aes(x = trialDuration, y = changepoints, colour = as.factor(targetCollected))) + 
geom_point() +
scale_color_brewer(palette = "Set2") +
facet_wrap(~ studyGroup, ncol=6) + 
theme_minimal() + 
theme(legend.position = "bottom") + 
labs(x = "trialDuration", y = "changepoints", colour = "targetCollected")
```

# Correlations for factor(studyGroup)+changepoints+straightness.index+trialDuration+interactions
```{r}
cor_dataset_SpatMem <- dplyr::select(dataset_SpatMem_1a, changepoints, straightness.index, trialDuration, interactions, trial_count, studyGroup)

par(mfrow=c(2,3))
for (i in c("Bantu_Child", "Mbendjele_Child", "US_Adult", "US_Child", "bonobo", "chimp")) {
  temp <- subset(cor_dataset_SpatMem, studyGroup == i)[,1:ncol(cor_dataset_SpatMem)-1]
  corr <- cor(temp, use = "complete.obs")
  corrplot(corr, method = "ellipse", title = paste0("Correlation for ", i), cex.main = 0.8, tl.cex = 0.8, tl.col = 'black', mar=c(0.5, 0.5, 1, 0.5), type = "upper")
  }
```

# Correlation values in a facet grid for changepoints and straightness.index/trialDuration group by studyGroup across targetCollected
```{r}
ggplot(dataset_SpatMem_1a, aes(changepoints, straightness.index)) + 
  geom_smooth(method = "loess", colour = "red", fill = "red") + 
  geom_smooth(method = "lm", colour = "blue", fill = "blue") + 
  ylim(0,2) +
  geom_point(size=.5) + 
  facet_grid(targetCollected~studyGroup, margins=TRUE)

ggplot(dataset_SpatMem_1a, aes(changepoints, straightness.index)) + 
  geom_smooth(method = "lm") + 
  ylim(0,2) +
  geom_point(size=.5) + 
  facet_grid(targetCollected~studyGroup, margins=TRUE)
```

# Boxplot facet
```{r}
ggplot(dataset_SpatMem_1a, aes(x=straightness.index, y=changepoints, group=studyGroup)) + 
  geom_boxplot(aes(fill=studyGroup)) +
  facet_grid(targetCollected~studyGroup, margins=TRUE)

ggplot(dataset_SpatMem_1a, aes(x=trialDuration, y=changepoints, group=studyGroup)) + 
  geom_boxplot(aes(fill=studyGroup)) +
  facet_grid(targetCollected~studyGroup, margins=TRUE) +
  theme(axis.text.x = element_text(angle =45, hjust = 1))
```

# Try PCA for targetCollected/changepoints/straightness.index/trialDuration/interactions/trial_count/studyGroup
```{r}
SpatMem_4 <- na.omit(dplyr::select(dataset_SpatMem_1a, targetCollected, changepoints, straightness.index, trialDuration, interactions, trial_count, studyGroup))
  
# Prepare the data
SpatMem_4_pca <- na.omit(SpatMem_4[,c("targetCollected","changepoints","interactions","trial_count","straightness.index","trialDuration")])
# Principal component analysis
pca <- prcomp(SpatMem_4_pca, scale. = TRUE)
# Plot
autoplot(pca, loadings = TRUE, loadings.label = TRUE,
         data = SpatMem_4, colour = 'studyGroup', xlim = c(-0.15, 0.1))
```

# ???Plotting cluster 
```{r}
autoplot(clara(SpatMem_4[,c("targetCollected","changepoints","interactions","trial_count","straightness.index","trialDuration")], 3))
autoplot(fanny(SpatMem_4[,c("targetCollected","changepoints","interactions","trial_count","straightness.index","trialDuration")], 3), frame = TRUE)
autoplot(pam(SpatMem_4[,c("targetCollected","changepoints","interactions","trial_count","straightness.index","trialDuration")], 3), frame = TRUE, frame.type = 'norm')
autoplot(silhouette(pam(SpatMem_4[,c("targetCollected","changepoints","interactions","trial_count","straightness.index","trialDuration")], 3L)))
```

# ???Plotting K-means
```{r}
autoplot(kmeans(SpatMem_4[,c("targetCollected","changepoints","interactions","trial_count","straightness.index","trialDuration")], 3), data = SpatMem_4[,c("targetCollected","changepoints","interactions","trial_count","straightness.index","trialDuration")], label = TRUE, label.size = 3, frame = TRUE)
```

# Semi-supervised Local Fisher Discriminant Analysis (SELF)
```{r}
model <- self(SpatMem_4[,c("targetCollected","changepoints","interactions","trial_count","straightness.index","trialDuration")], SpatMem_4[, "studyGroup"], beta = 0.1, r = 3, metric="plain")
autoplot(model, data = SpatMem_4, frame = TRUE, frame.colour = 'studyGroup')
```

# Try Lasso for targetCollected/changepoints/interactions/trial_count/studyGroup - straightness.index/trialDuration
```{r}
SpatMem_4 <- na.omit(dplyr::select(dataset_SpatMem_1a, targetCollected, changepoints, straightness.index, trialDuration, interactions, trial_count, studyGroup))
```

```{r}
for (i in c("Bantu_Child", "Mbendjele_Child", "US_Adult", "US_Child", "bonobo", "chimp")) {
  temp <- SpatMem_4 %>% filter(studyGroup == i)
  y_nom <- temp$straightness.index
  x_nom <- data.matrix(temp[, c("targetCollected","changepoints","interactions","trial_count")])
  
  #perform k-fold cross-validation to find optimal lambda value
  set.seed(2023)
  cv_model <- cv.glmnet(x_nom, y_nom, alpha = 1)
  
  #find coefficients of best model
  best_model <- glmnet(x_nom, y_nom, alpha = 1, lambda = cv_model$lambda.min)
  print(coef(best_model))

  best_model2 <- glmnet(x_nom, y_nom, alpha = 1, lambda = cv_model$lambda.1se)
  print(coef(best_model2))
}
  
#SpatMem_4_Bantu_Child <- SpatMem_4 %>% filter(studyGroup == "Bantu_Child")
#SpatMem_4_Mbendjele_Child <- SpatMem_4 %>% filter(studyGroup == "Mbendjele_Child")
#SpatMem_4_US_Adult <- SpatMem_4 %>% filter(studyGroup == "US_Adult")
#SpatMem_4_US_Child <- SpatMem_4 %>% filter(studyGroup == "US_Child")
#SpatMem_4_bonobo <- SpatMem_4 %>% filter(studyGroup == "bonobo")
#SpatMem_4_chimp <- SpatMem_4 %>% filter(studyGroup == "chimp")
```

```{r}
for (i in c("Bantu_Child", "Mbendjele_Child", "US_Adult", "US_Child", "bonobo", "chimp")) {
  temp <- SpatMem_4 %>% filter(studyGroup == i)
  y_nom <- temp$trialDuration
  x_nom <- data.matrix(temp[, c("targetCollected","changepoints","interactions","trial_count")])
  
  #perform k-fold cross-validation to find optimal lambda value
  set.seed(2023)
  cv_model <- cv.glmnet(x_nom, y_nom, alpha = 1)
  
  #find coefficients of best model
  best_model <- glmnet(x_nom, y_nom, alpha = 1, lambda = cv_model$lambda.min)
  print(coef(best_model))

  best_model2 <- glmnet(x_nom, y_nom, alpha = 1, lambda = cv_model$lambda.1se)
  print(coef(best_model2))
}
```


